---
title: "Custom neighborhoods in the raceland package"
author: Jakub Nowosad
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

A long, long time ago (precisely speaking -- on 17th of June, 2021) I met with Tomek on Skype.
We discussed various topics, with one being the ability to specify different neighborhoods (types of cell adjacencies) for the IT-metrics calculations in **raceland**.
Finally, I found some time to look at the code and added this option. 
This document shows how to use different neighborhoods in **raceland**.^[This feature requires the terra branch of the raceland package to be installed - `remotes::install_github("nowosad/raceland@terra")`.]

```{r}
# remotes::install_github("nowosad/raceland@terra")
```

For this purpose, we will use example data available in the package.

```{r}
library(raceland)
library(terra)
library(dplyr)
library(tidyr)
library(sf)
library(stringr)
library(ggplot2)
theme_set(theme_bw())
race_raster = rast(system.file("extdata/race_raster.tif", package = "raceland"))
list_raster = list.files(system.file("rast_data", package = "raceland"),
                         full.names = TRUE)
race_raster = rast(list_raster)
plot(race_raster)
```

The first two steps are the same as in the original workflow: 1) Creating realizations; 2) Calculating local subpopulation densities.

```{r}
set.seed(2022-02-14)
# constructing racial landscape
real_raster = create_realizations(x = race_raster, n = 100)
# calculating local subpopulation densities
dens_raster = create_densities(real_raster, race_raster, window_size = 10)
```

Next, in the original workflow, we would calculate IT-derived metrics, based on cell adjacencies in a 4- or 8-neighborhood.

```{r}
metr_df_20 = calculate_metrics(x = real_raster, w = dens_raster, 
                               neighbourhood = 4, fun = "mean", 
                               size = 20, threshold = 0.5)
```

Now, however, we are able to define any neighborhood.
This can be done by creating a neighborhood matrix, which should have one cell with value 0 (the focal cell), and at least one cell with value 1 (the adjacent cells).
Cells with other values (e.g., NA) are ignored.

```{r}
create_neigh = function(size){
  r = matrix(nrow = size, ncol = size)
  cents = c(seq(1, ncol(r), by = 2), seq(ncol(r), 1, by = -2)[-1])
  inrow = c(seq(ncol(r), 1, by = -2), seq(1, ncol(r), by = 2)[-1])
  for (i in 1:nrow(r)){
    r[i, ] = c(rep(NA, (inrow[i] - 1)/2), rep(1, cents[i]), rep(NA, (inrow[i] - 1)/2))
  }
  r[ceiling(length(r)/2)] = 0
  return(r)
}
```

For example, see four custom neighborhood matrices below.

```{r, fig.asp=0.25}
par(mfcol = c(1, 4))
terra::plot(terra::rast(create_neigh(3), extent = terra::ext(0, 1, 0, 1)), main = 3, mar = c(0,1,2,0), axes = FALSE)
terra::plot(terra::rast(create_neigh(5), extent = terra::ext(0, 1, 0, 1)), main = 5, mar = c(0,1,2,0), axes = FALSE)
terra::plot(terra::rast(create_neigh(7), extent = terra::ext(0, 1, 0, 1)), main = 7, mar = c(0,1,2,0), axes = FALSE)
terra::plot(terra::rast(create_neigh(9), extent = terra::ext(0, 1, 0, 1)), main = 9, mar = c(0,1,2,0), axes = FALSE)
```

Each of the neighborhood matrices shown above can be used to define the number of directions in which cell adjacencies are considered as neighbors for IT metrics calculations.
The rest of the document shows the ENT and MUTINF values for different areas (rows/cols) and different neighborhoods.

```{r}
metr_df_20_3 = calculate_metrics(x = real_raster, w = dens_raster, 
                               neighbourhood = create_neigh(3), fun = "mean", 
                               size = 20, threshold = 0.5)
metr_df_20_5 = calculate_metrics(x = real_raster, w = dens_raster, 
                               neighbourhood = create_neigh(5), fun = "mean", 
                               size = 20, threshold = 0.5)
metr_df_20_7 = calculate_metrics(x = real_raster, w = dens_raster, 
                               neighbourhood = create_neigh(7), fun = "mean", 
                               size = 20, threshold = 0.5)
metr_df_20_9 = calculate_metrics(x = real_raster, w = dens_raster, 
                               neighbourhood = create_neigh(9), fun = "mean", 
                               size = 20, threshold = 0.5)
```


```{r, message=FALSE}
smr_20_3 = metr_df_20_3 %>%
  group_by(row, col) %>%
  summarize(ent_mean3 = mean(ent, na.rm = TRUE), mutinf_mean3 = mean(mutinf, na.rm = TRUE))
smr_20_5 = metr_df_20_5 %>%
  group_by(row, col) %>%
  summarize(ent_mean5 = mean(ent, na.rm = TRUE), mutinf_mean5 = mean(mutinf, na.rm = TRUE))
smr_20_7 = metr_df_20_7 %>%
  group_by(row, col) %>%
  summarize(ent_mean7 = mean(ent, na.rm = TRUE), mutinf_mean7 = mean(mutinf, na.rm = TRUE))
smr_20_9 = metr_df_20_9 %>%
  group_by(row, col) %>%
  summarize(ent_mean9 = mean(ent, na.rm = TRUE), mutinf_mean9 = mean(mutinf, na.rm = TRUE))
```


```{r, message=FALSE}
grid_sf20 = create_grid(real_raster, size = 20)
# join IT-metrics to the grid
attr_grid20 = left_join(grid_sf20, smr_20_3, by = c("row", "col")) |> 
  left_join(smr_20_5) |>
  left_join(smr_20_7) |>
  left_join(smr_20_9)
```


```{r}
attr_grid20_df = attr_grid20 |>
  st_drop_geometry() |>
  pivot_longer(cols = c(-row, -col)) |>
  mutate(size = str_extract(name, "[0-9]")) |>
  mutate(name = str_remove_all(name, "[0-9]"),
         rowcol = paste0(row, col))
```

```{r}
# ggplot(attr_grid20_df, aes(x = size, value, color = rowcol)) +
#   geom_point() +
#   facet_wrap(name~., scales = "free_y")
```

```{r, out.width="50%", fig.align="center"}
hex_colors = c("#F16667", "#6EBE44", "#7E69AF", "#C77213","#F8DF1D")
realization = create_realizations(race_raster, 1)
plot_realization(realization, race_raster, hex = hex_colors)
plot(attr_grid20, add = TRUE, col = NA)
```

```{r, fig.align="center"}
g1 = ggplot(subset(attr_grid20_df, name == "ent_mean"),
       aes(x = size, value)) +
  geom_point() +
  facet_grid(row~col) +
  labs(title = "ENT", y = NULL) +
  theme(aspect.ratio = 1)
g2 = ggplot(subset(attr_grid20_df, name == "mutinf_mean"),
       aes(x = size, value)) +
  geom_point() +
  facet_grid(row~col) +
  labs(title = "MUTINF", y = NULL) +
  theme(aspect.ratio = 1)
cowplot::plot_grid(g1, g2, nrow = 1)
```
